General purpose gene annotation using deep learning

### 2020-01-30 paper/main.tex
paper

\section{Approach}
Helixer is a prototype software for training and utilizing a general purpose gene annotation model for eukaryotic genomes. We demonstrate the effectiveness of this approach by training a single model for animals (phylogenetic names?) and plants respectively, which was then evaluated on 192 genomes in the animal case and 60 genomes for the plant model. We leverage modelling techniques from the field of deep learning, which has been shown to have the potential to be superior in flexibility and accuracy to traditional machine learning methods in a variety of fields. Our model expects the raw genomic sequence as main input and produces functional predictions on a base pair wise level. The source code and all data is publically available.

\begin{methods}
\section{Methods}
We worked with 192 animal and 60 plant genomes. The data of each genome consists of the latest publically available genomic assembly in form of a FASTA file and the latest annotation in the GFF format. We used all animals genomes in EnsemblMetazoa 45 (?) \citealp{Howe19} as well as those of all non-embargoed plant genomes of the JGI Phytosome 13 database (?, citation). Both data groups were used seperately during training as well as during evaluation. At the start of our work we split both genome groups into training and test sets. The test sets contained 19 or 6 genomes, respectively. The test genomes were chosen based on being frequently used model species while also representing a broad phylogenetic spread. Some of the remaining genomes were used for training purposes and all were used for evaluating our models.

We pre-processed and stored the genomic information by using Geenuff \citealp{Denton19}. Geenuff is a tool for checking and exploring genomic data, which stores all information inside a SQL database. Training and evaluation data was generated by querying this database and transforming the returned data into a numerical format suitable for machine learning. In the case of unambiguous input data, the genomic sequence was transformed into matrix that represents a one hot encoding with 4 rows. For the other case see Suplementary material .. (). The annotations are also transformed into a 4-row one hot encoding matrix, where a one in each of the rows represents one of the possible classes {\it Intergenic}, {\it UTR}, {\it CDS} or {\it Intron}. An non-coding intron can be inferred from the prediction order. A full description of all generated data arrays can be found in table \ref{Tab:01}.

@@

We use metrics two metrics to judge model performance. The {\it Genic F1} is our primary metric that provides the most comprehensive picture of annotation quality in one number. The {\it Subgenic F1} is similar to the Genic F1, except that it does not take UTR predictions into account. This is done for comparability with Augustus and further explained in Chapter \ref{augustus}. Both metrics work by summing up True Positives (TP), False Positives (FP) and False Negatives (FN) of each considered class before calculating precision and recall from these values and combining the two into a F1 score. Both metrics do not take the values of the intergenic class into account, as this class is very abundant and appears to be by far the easiest to predict. In the case of the Genic F1, this means that effectively only the TP of the intergenic class are disregarded. The metrics essentially provide a weighted mean of the performances in the considered classes. All individual class performances are reported in Suplementary Material (?).

@@

The model was trained with 10 bases of genomic sequence as input during each time step and produces individual predictions for each of those 10 bases simultaneously. This grouping enabled us to train with far longer sequences than usual. A class wise evaluation including the calculation of the Genic F1 score is performed for the validation set after every epoch. We use the model with the best Genic F1 score unless the improvement to the next best one was less than $0.01$ apart in order to prevent overfitting.

	We also used multiple techniques to improve the prediction quality after the training was done. One very effective way was to input longer sequences during inference than during training. The sequences were up to 10 times larger, depending on species group and assembly quality. We also reconstructed the final predictions of a single model from overlapping predictions that were cut to a core sequence. This was done to drastically reduce a typical bias of the models towards the beginning and end of each sequence (see figure \ref{fig:length_wise_bias}) and also to provide the model with multiple different starting points. The different overlapping sequences were recombined by averaging the softmax values of each base. This was also done when we finally ensembled the whole output of different models on the softmax value level from (8?) different models. For more details on these post-processing techniques see Suplementary Material (?).

>>>>

\section{Approach}
Helixer is a prototype software for training and utilizing a general purpose gene annotation model for large eukaryotic genomes. We demonstrate the effectiveness of this approach by training a single model for animals (phylogenetic names?) and plants respectively, which was then evaluated on 192 genomes in the animal case and 60 genomes for the plant model. We leverage modelling techniques from the field of deep learning, which has been shown to have the potential to be superior in flexibility and accuracy to traditional machine learning methods in a variety of fields. Our model expects the raw genomic sequence as main input and produces functional predictions on a base pair wise level. The source code and all data is publically available.

\begin{methods}
\section{Methods}
We worked with 192 animal and 60 plant genomes. The data of each genome consists of the latest publically available genomic assembly in form of a FASTA file and the latest annotation in the GFF format. We used all animals genomes in EnsemblMetazoa 45 (?) \citealp{Howe19} as well as those of all non-embargoed plant genomes of the JGI Phytosome 13 database (?, citation). Both data groups were used seperately during training as well as during evaluation. At the start of our work we split both genome groups into training and test sets. The test sets contained 19 or 6 genomes, respectively. The test genomes were chosen to be of the best possible quality (?) based on collected metadata while also representing a broad phylogenetic spread. Some of the remaining genomes were used for training purposes and all were used for evaluating our models.

We pre-processed and stored the genomic information by using Geenuff \citealp{Denton19}. Geenuff is a tool for checking and exploring genomic data, which stores all information inside a SQL database. Training and evaluation data was generated by querying this database and transforming the returned data into a numerical format suitable for machine learning. In the case of unambiguous input data, the genomic sequence was transformed into matrix that represents a one hot encoding with 4 rows. For the other case see Supplementary material .. (). The annotations are also transformed into a 4-row one hot encoding matrix, where a one in each of the rows represents one of the possible classes {\it Intergenic}, {\it UTR}, {\it CDS} or {\it Intron}. An non-coding intron can be inferred from the prediction order. A full description of all generated data arrays can be found in table \ref{Tab:01}.

@@

We use metrics two metrics to judge model performance. The {\it Genic F1} is our primary metric that provides the most comprehensive picture of annotation quality in one number. The {\it Subgenic F1} is similar to the Genic F1, except that it does not take UTR predictions into account. This is done for comparability with Augustus and further explained in Chapter \ref{augustus}. Both metrics work by summing up True Positives (TP), False Positives (FP) and False Negatives (FN) of each considered class before calculating precision and recall from these values and combining the two into a F1 score. Both metrics do not take the values of the intergenic class into account, as this class is very abundant and appears to be by far the easiest to predict. In the case of the Genic F1, this means that effectively only the TP of the intergenic class are disregarded. The metrics essentially provide a weighted mean of the performances in the considered classes. All individual class performances are reported in Supplementary Material (?).

@@

The model was trained with 10 bases of genomic sequence as input during each time step and produces individual predictions for each of those 10 bases simultaneously. This grouping enabled us to train effectively with far longer sequences than usual. A class wise evaluation including the calculation of the Genic F1 score is performed on the validation set after every epoch.

We also used multiple techniques to improve the prediction quality after the training was done. One very effective way for genomes with larger genes was to input longer sequences during inference than during training. This was done for all animal genomes except the invertebrates. The input sequences were up to 10 times larger, depending on species group and assembly quality. This also demonstrates the generalization capability of the model as it is able to make successful prediction on far longer sequences than it has ever seen. The sequence length during inference can be any of 20.000, 50.000, 100.000 or 200.000 and is determined by the species family(?). Families that tend to have longer genes get assigned longer sequence chunks. If the N75 of a species is less than twice as high as the supposed sequence length it is lowered until either criteria is met or a length of 50.000 is reached. For implementation details on this and the following techniques see Supplementary Material (?) or the source code..

We also construct the final predictions of a single model by overlapping predictions that were made from a sliding window and then cut to a core sequence. This was done to strongly reduce a typical drop in performance of the models towards the beginning and end of each sequence (see figure \ref{fig:length_wise_bias}) and also to improve model performance in general by providing the model with multiple different starting points. The different overlapping sequences were recombined by averaging the individual softmax values of each base.

A model ensemble with 8 components was used to generate the final predictions for each species. For this, we trained 4 seperate animal and plant models and also used a model from an epoch that did not have the highest Genic F1 on the validation set. Depending on the results, we used a model from an epoch with either the highest genic recall or precision. The model with the best Genic F1 was always best in one of those. This was done to increase the diversity of the model ensemble. As with the overlapping, the fusion of the 8 individual predictions was done by a averaging the softmax values of each base pair prediction.
